Visualization and analysis of probability distributions of large temporal data

### 2021-11-15 Rmd/04-gracsr.Rmd
ch: to ch-
paper-gracsr update

# Clustering time series based on probability distributions across temporal granularities {#ch:gracsr}

With more and more time series data being collected at much finer temporal resolution, for a longer length of time, and for a larger number of individuals/entities, time series clustering research is getting a lot of traction. Long, noisy, patchy, uneven, and asynchronous time series are common in many fields, limiting similarity searches or lowering method efficiency when clustering is based on a distance metric. In this work, we suggest two approaches for obtaining similarity between time series based on probability distributions over cyclic temporal granularities for distance-based clustering approaches. Cyclic granularities like hour-of-the-day, work-day/weekend, month-of-the-year and so on, are useful for finding repeated patterns in the data. Looking at probability distributions across cyclic granularities serves two purposes: (a) "Probability distributions" characterize the inherent temporal data structure of these large unequal-length time series and are robust to missing or noisy data. (b) Using probability distributions over "cyclic granularities" ensures small pockets of similar "repeated" behaviors. Our method is capable of producing useful clusters, as demonstrated by testing on validation data designs and a sample of residential smart meter consumers.

```{r external3, include = FALSE, cache = FALSE}
knitr::read_chunk('scripts/gracsr.R')
```

@@

@@

<!-- time series clustering and its challenges -->
Time-series clustering is the process of unsupervised partitioning of $n$ time-series data into $k$ ($k<n$) meaningful groups such that homogeneous time-series are grouped together based on a certain similarity measure. The time-series features, length of time-series, representation technique, and, of course, the purpose of clustering time-series all influence the suitable similarity measure or distance metric to a meaningful level. The three primary methods to time series clustering (@liao2005clustering) are algorithms that operate directly with distances or raw data points in the time or frequency domain (distance-based), with features derived from raw data (feature-based), or indirectly with models constructed from raw data (model-based). The efficacy of distance-based techniques is highly dependent on the distance measure utilized. Defining an appropriate distance measure for the raw time series may be a difficult task since it must take into account noise, variable lengths of time series, asynchronous time series, different scales, and missing data. Commonly used Distance-based similarity measures as suggested by a decade review of time series clustering approaches (@Aghabozorgi2015-ct) are Euclidean, Pearson's correlation coefficient and related distances, Dynamic Time Warping, Autocorrelation, Short time series distance, Piecewise regularization, cross-correlation between time series, or a symmetric version of the Kullback–Liebler distances (@liao2007clustering) but on a vector time series data. Among these alternatives, Euclidean distances have high performance but need the same length of data over the same period, resulting in information loss regardless of whether it is on raw data or a smaller collection of features. DTW works well with time series of different lengths (@corradini2001dynamic), but it is incapable of handling missing observations. Surprisingly, probability distributions, which may reflect the inherent temporal structure of a time series have not been considered in determining time series similarity.

\noindent We consider the problem of clustering a large number of univariate time series of continuous values which are available at fine temporal scales, being motivated by the residential smart meter data. These time series data are long (with more and more data collected at finer resolutions), are asynchronous, with varying time lengths for different houses and missing observations and characterized by noisy and patchy behavior that can quickly become overwhelming and hard to interpret, requiring summarizing the large number of customers into pockets of similar energy behavior. Choosing probability distributions seem to be a natural way to analyze these types of data sets since they are robust to uneven length, missing data, or noise. This paper proposes two approaches for obtaining pairwise similarities based on Jensen-Shannon distances between probability distributions across significant cyclic granularities. Cyclic temporal granularities [@Gupta2021-hakear], which are temporal deconstructions of a time period into units such as hour-of-the-day, work-day/weekend, can be useful for measuring repetitive patterns in large univariate time series data. The resulting clusters are expected to group customers that have similar repetitive behaviors across each of the interesting cyclic granularities. Below are some of the benefits of our method:

- When using probability distributions, data does not have to be the same length or observed during the exact same time period (unless there is a structural pattern);

@@

- Jensen-Shannon distances evaluate the distance between two distributions rather than raw data, which is less sensitive to missing observations and outliers than other conventional distance methods;

- While most clustering algorithms produce clusters similar across just one temporal granularity, this technique takes a broader approach to the problem, attempting to group observations with similar distributions across all interesting cyclic granularities;

@@

_Background and motivation_

Large spatio-temporal data sets, both from open and administrative sources, offer up a world of possibilities for research. One such data sets for Australia is the Smart Grid, Smart City (SGSC) project (2010–2014) available through [Department of the Environment and Energy](https://data.gov.au/data/organization/doee) The project provides half-hourly data of over 13,000 household electricity smart meters distributed unevenly from October 2011 to March 2014. Larger data sets include greater uncertainty about customer behavior due to growing variety of customers. Households vary in size, location, and amenities such as solar panels, central heating, and air conditioning. The behavioral patterns differ amongst customers due to many temporal dependencies. Some households use a dryer, while others dry their clothes on a line. Their weekly profile may reflect this. They may vary monthly, with some customers using more air conditioners or heaters than others, while having equivalent electrical equipment and weather circumstances. Some customers are night owls, while others are morning larks. Day-off energy use varies depending on whether customers stay home or go outside. Age, lifestyle, family composition, building attributes, weather, availability of diverse electrical equipment, among other factors, make the task of properly segmenting customers into comparable energy behavior a fascinating one. This challenge is worsened when all we know about our consumers is their energy use history (@Ushakova2020-rl). To safeguard the customers' privacy, it is probable that such information is not accessible. Also, energy suppliers may not always update client information, such as property features, in a timely manner. Thus, there is a growing need to have research that examines how much energy usage heterogeneity can be found in smart meter data and what are some of the most common power consumption patterns, rather than explaining why consumption differs.

_Related work_

A multitude of papers have emerged around smart meter time series clustering for deepening our knowledge of consumption patterns. @Tureczek2017-pb conducted a systematic study of over $2100$ peer-reviewed papers on smart meter data analytics. None of the $34$ articles chosen for their emphasis use Australian smart meter data. The most often used algorithm is K-Means. Using K-Means without considering time series structure or correlation results in inefficient clusters. Principal Component Analysis (PCA) or Self-Organizing Maps (SOM) eliminate correlation patterns and decrease feature space, but lose interpretability. To reduce dimensionality, several studies use principal component analysis or factor analysis to pre-process smart-meter data before clustering (@Ndiaye2011-pf). Other algorithms utilized in the literature include k-means variants, hierarchical approaches, and greedy k-medoids. Time series data, such as smart meter data, are not well-suited to any of the techniques mentioned in @Tureczek2017-pb. Only one study [@ozawa2016determining] identified time series characteristics using Fourier transformation, which converts data from time to frequency and then uses K-Means to cluster by greatest frequency. @Motlagh2019-yj suggests that the time feature extraction is limited by the type of noisy, patchy, and unequal time-series common in residential and addresses model-based clustering by transforming the series into other other objects such as structure or set of parameters which can be more easily characterized and clustered. [@chicco2010renyi] addresses information theory-based clustering such as Shannon or Renyi entropy and its variations. @Melnykov2013-sp discusses how outliers, noisy observations and scattered observations can complicate estimating mixture model parameters and hence the partitions. To our knowledge, none of the methods focus on exploring heterogeneity in repetitive behaviors based on the dynamics of multiple temporal dependencies using probability distributions.

@@

The remainder of the paper is organized as follows: Section&nbsp;\ref{sec:methodology} provides the clustering methodology. Section&nbsp;\ref{sec:validation} shows data designs to validate our methods. Section&nbsp;\ref{sec:cases} discusses the application of the method to a subset of the real data. Finally, we summarize our results and discuss possible future directions in Section&nbsp;\ref{sec:conclusion}.

@@

The foundation of our method is unsupervised clustering algorithms based exclusively on the time-series data. The proposed methodology aims to leverage the intrinsic data structure hidden within cyclic temporal granularities. The existing work on clustering probability distributions assumes we have an iid sample $f_1(v),\dots,f_n(v)$, where $f_i(v)$ denotes the distribution from observation $i$ over some random variable $v = \{v_t: t = 0, 1, 2, \dots, T-1\}$ observed across $T$ time points. <!--So going back to the smart meter example, $f_i(v)$ is the distribution of customer $i$ and $v$ is electricity demand.--> In this work, instead of considering the probability distributions of the linear time series, we assume it across different categories of any cyclic granularity. We can consider categories of an individual cyclic granularity ($A$) or combination of categories for two interacting granularities ($A*B$) to have a distribution, where $A, B$ are defined as $A = \{a_j: j=1, 2, \dots J\}$ and $B = \{b_k: k = 1, 2, \dots K\}$. For example, let us consider two cyclic granularities $A$ and $B$ representing hour-of-day and day-of-week. Then $A = \{0, 1, 2, \dots, 23\}$ and $B = \{Mon, Tue, Wed, \dots, Sun\}$. In case individual granularities are considered, there are $J = 24$ distributions of the form $f_{i,j}(v)$ or $K = 7$ distributions of the form $f_{i,k}(v)$ for each customer $i$. In case of interaction, $J*K=168$ distributions of the form $f_{i,j, k}(v)$ could be conceived for each customer $i$. Hence clustering these customers is equivalent to clustering these collections of conditional univariate probability distributions. Towards this goal, we need to decide how to measure similarities between collections of univariate probability distributions. There are multiple ways to measure similarities depending on the aim of the analysis. <!--This paper focuses on looking at the (dis) similarity between underlying distributions that may have resulted in different patterns across different cyclic temporal granularities, that eventually have resulted in the (dis) similarity between time series. --> This paper considers a two approaches for finding distances between time series. Both of these approaches may be useful in a practical context and, depending on the data set, may or may not propose the same customer classification. The obtained distances could be fed into a clustering algorithm to break large data sets into subgroups that can then be analyzed separately. <!--These clusters may be commonly associated with real-world data segmentation. However, since the data is unlabeled a priori, more information is required to corroborate this.--> The methodology is explained in the Figure \ref{fig:flowchart} and each element of the pipeline is discussed.

```{r flowchart, fig.cap = "Flow chart illustrating the pipeline for methodology", out.width="100%"}
knitr::include_graphics("img/flowchart.png")

@@

- _Find significant granularities or harmonies_

[@Gupta2021-hakear] proposes a method for choosing significant cyclic granularities and harmonies (interacting granularities that can be studied together) [@Gupta2021-gravitas], which is used in this work. We define "significant" granularities as those with significant distributional differences across categories. It is preferable to consider those since it is more probable that there will be some intriguing repeated behavior worth investigating. It should be noted that all of the observations in the study may not have the same set of important granularities. The following is a method for generating a list ($S_c$) of significant granularities for all observations:

 (a) Remove granularities from the comprehensive list that are insignificant for all observations.

 (b) select only the granularities that are significant for the majority of observations.

There will be observations in both cases where one or a few selected granularities are boring. Even in that case, having this group of observations with no interesting patterns at a granularity that regularly discovers patterns may be valuable. If, on the other hand, the granularities under examination are truly important for a group of data, unique patterns may be identified while clustering them.

- _Data transformation_

Observations often have a somewhat skewed time series distribution and their ranges might vary greatly. Statistical transformations are employed to bring all of them to the same range or normalize each observation. This is important because we are not interested in trivial clusters that vary in magnitudes, but rather in uncovering comparable patterns of distributional differences between categories. For the JS-based approaches, two data transformation techniques are utilized viz, Normal-Quantile Transform (NQT) and Robust scaling. NQT is a built-in transformation for computing $wpd$, which is the foundation of wpd-based distances.

_Robust scaling_ The normalized $i^{th}$ observation is denoted by $v_{norm} = \frac{v_t - q_{0.5}}{q_{0.75}-q_{0.25}}$, where $v_t$ is the actual value at the $t^{th}$ time point and $q_{0.25}$, $q_{0.5}$ and $q_{0.75}$ are the $25^{th}$, $50^{th}$ and $75^{th}$ percentile of the time series for the $i^{th}$ observation. $v_{norm}$ has zero mean and median, as well as a standard deviation of one, with the outliers having same relative connections to other values.

_Normal-Quantile transform_ The raw data for all observations is individually normal-quantile transformed (NQT) [@Krzysztofowicz1997-bv], so that the transformed data follows a standard normal distribution. NQT will make the skewed distributions bell-shaped. As a result, determining which raw distribution was used is difficult using the modified distribution. Multimodality is also concealed or reversed. As a result, while displaying transformed data using NQT, one must exercise caution. But NQT is a useful transformation that often improves clustering performance.

@@

- _Data pre-preprocessing_

We start with a "tsibble" (@wang2020tsibble) data structure with an index variable representing inherent ordering from past to present and a key variable that defines observational units over time. The measured variable for an observation is a time-indexed sequence of values. This sequence, however, could be shown in several ways. A shuffle of the raw sequence may represent hourly consumption throughout a day, a week, or a year. Cyclic granularities can be expressed in terms of the index set in the "tsibble" data structure. But the data structure changes while transporting from linear to cyclic scale of time as multiple observations now correspond to each category and induce a probability distribution. Directly computing Jensen-Shannon distances between the entire probability distributions can be very costly. Hence, in this paper, quantiles are chosen to characterize the probability distributions. So, in the final data structure, each category of a cyclic granularity corresponds to a list of numbers which is essentially a few chosen quantiles.

- _Distance metrics_

@@

The total (dis) similarity between each pair of customers is obtained by combining the distances between the collections of conditional distributions. This needs to be done in a way such that the resulting metric is a distance metric, and could be fed into the clustering algorithm. Two types of distance metric is considered:

@@

**JS-based distances**

@@

\noindent The sum of the distances between two observations $x$ and $y$ in terms of cyclic granularity $A$ is defined as $$S^A_{x,y} = \sum_{j} D_{x,y}(A)$$ (sum of distances between each category $j$ of cyclic granularity $A$) or $$S^{A*B}_{x,y} = \sum_j \sum_k D_{x,y}(A, B)$$ (sum of distances between each combination of categories $(j, k)$ of the harmony $(A, B)$. After determining the distance between two series in terms of one granularity, we must combine them to produce a distance based on all significant granularities. When combining distances from individual $L$ cyclic granularities $C_l$ with $n_l$ levels, $$S_{x, y} = \sum_lS^{C_l}_{x,y}/n_l$$ is employed, which is also a distance metric since it is the sum of JS distances. This approach is expected to yield groups, such that the variation in observations within each group is in magnitude rather than distributional pattern, while the variation between groups is only in distributional pattern across categories.

**wpd-based distances**

Compute weighted pairwise distances $wpd$ (@Gupta2021-hakear) for all considered granularities for all observations. $wpd$ is designed to capture the maximum variation in the measured variable explained by an individual cyclic granularity or their interaction and is estimated by the maximum pairwise distances between consecutive categories normalized by appropriate parameters. A higher value of $wpd$ indicates that some interesting pattern is expected, whereas a lower value would indicate otherwise.

\noindent Once we have chosen $wpd$ as a relevant feature for characterizing the distributions across one cyclic granularity, we have to decide how we combine differences between the multiple features (corresponding to multiple granularities) into a single number. The Euclidean distance between them is chosen, with the granularities acting as variables and $wpd$ representing the value under each variable. With this approach, we should expect the observations with similar $wpd$ values to be clustered together. Thus, this approach is useful for grouping observations that have a similar significance of patterns across different granularities. Similar significance does not imply a similar pattern, which is where this technique varies from JS-based distances, which detect differences in patterns across categories.

@@

- _Clustering algorithm_

@@

- _Characterization of clusters_

\noindent Cluster characterization is an important element of cluster analysis. @Cook2007-qe provides several methods for characterizing clusters. _Parallel coordinate plots_ (@wegman1990hyperdimensional), _Scatterplot matrix_, _Displaying cluster statistics_ (@dasu2005grouping), _MDS_ (@borg2005modern), PCA, _t-SNE_(@R-tsne), _Tour_ (@wickham2011tourr) are some of the graphical approaches used in this study. A Parallel Coordinates Plot features parallel axes for each variable and each axis is linked by lines. Changing the axes may reveal patterns or relationships between variables for categorical variables. However, for categories with cyclic temporal granularities, preserving the underlying ordering of time is more desirable. Displaying cluster statistics is useful when we have larger problems and it is difficult to read the parallel coordinate plots due to congestion. All of MDS, PCA and t-SNE use a distance or dissimilarity matrix to construct a reduced-dimension space representation, their goals are diverse. Multidimensional scaling (@borg2005modern) seeks to maintain the distances between pairs of data points, with an emphasis on pairings of distant points in the original space. The t-SNE embedding will compress data points that are close in high-dimensional space. Tour is a collection of interpolated linear projections of multivariate data into lower-dimensional space. The cluster characterization approach varies depending on the distance metric used. Parallel coordinate plots, scatter plot matrices, MDS or PCA are potentially useful ways to characterize clusters using wpd-based distances. For JS-based distances, plotting cluster statistics is beneficial for characterization and variable importance could be displayed through parallel coordinate plots.

@@

### Data generating processes

@@

_Scenario (a): All significant granularities_

@@

\noindent Consider the instance where $g1$, $g2$, and $g3$ all contribute to design distinction. This means that each granularity will have significantly different patterns at least across one of the designs to be clustered. In Table \ref{tab:tab-dist-design} various distributions across categories are considered (top) which lead to different designs (bottom). Figure \ref{fig:plot-3gran-new} shows the simulated variable's linear (left) and cyclic (right) representations for each of these five designs. The structural difference in the time series variable is impossible to discern from the linear view, with all of them looking very similar. The shift in structure may be seen clearly in the distribution of cyclic granularities. The following scenarios use solely graphical displays across cyclic granularities to highlight distributional differences in categories.

>>>>

# Clustering time series based on probability distributions across temporal granularities {#ch-gracsr}

Clustering is a potential approach for Organizing large collections of times series into small homogeneous groups, but a difficult step is determining an appropriate metric to measure similarity between time series. The similarity metric needs to be capable of accommodating long, noisy, and asynchronous time series and also capture cyclical patterns. In this paper, two approaches for measuring distances between time series are presented, based on probability distributions over cyclic temporal granularities. Both are compatible with a variety of clustering algorithms. Cyclic granularities like hour-of-the-day, work-day/weekend, month-of-the-year, are useful for finding repeated patterns in the data. Measuring similarity based on probability distributions across cyclic granularities serves two purposes: (a) characterizing the inherent temporal data structure of long, unequal-length time series in a manner robust to missing or noisy data; (b) small pockets of similar repeated behaviors can be captured. This approach is capable of producing useful clusters, as demonstrated on validation data designs and a sample of residential smart meter records.

```{r external-gracsr, include = FALSE, cache = FALSE}
#knitr::read_chunk('script/main.R')
knitr::read_chunk('scripts/gracsr/mylibrary.R')
knitr::read_chunk('scripts/gracsr/mytheme.R')
knitr::read_chunk('scripts/gracsr/application.R')
knitr::read_chunk('scripts/gracsr/application.R')
knitr::read_chunk('scripts/gracsr/validation.R')
```

@@

```{r mytheme-application2}

```{r mytheme-application3}

@@

<!-- time series clustering and its challenges -->
Time-series clustering is the process of unsupervised partitioning of $n$ time-series data into $k$ ($k<n$) meaningful groups such that homogeneous time-series are grouped together based on a certain similarity measure. The time-series features, length of time-series, representation technique, and, of course, the purpose of clustering time-series all influence the suitable similarity measure or distance metric to a meaningful level. The three primary methods to time series clustering (@liao2005clustering) are algorithms that operate directly with distances or raw data points in the time or frequency domain (distance-based), with features derived from raw data (feature-based), or indirectly with models constructed from raw data (model-based). The efficacy of distance-based techniques is highly dependent on the distance measure utilized. Defining an appropriate distance measure for the raw time series may be a difficult task since it must take into account noise, variable lengths of time series, asynchronous time series, different scales, and missing data. Commonly used distance-based similarity measures as suggested by a decade review of time series clustering approaches (@Aghabozorgi2015-ct) are Euclidean, Pearson's correlation coefficient and related distances, Dynamic Time Warping (DTW), Autocorrelation, Short time series distance, Piecewise regularization, cross-correlation between time series, or a symmetric version of the Kullback–Liebler distances (@liao2007clustering) but on a vector time series data. Among these alternatives, Euclidean distances have high performance but need the same length of data over the same period, resulting in information loss regardless of whether it is on raw data or a smaller collection of features. DTW works well with time series of different lengths (@corradini2001dynamic), but it is incapable of handling missing observations. Surprisingly, probability distributions, which may reflect the inherent temporal structure of a time series have not been considered in determining time series similarity.

\noindent This work is motivated by a need to cluster a large collection of residential smart meter data, so that households can be grouped into similar energy usage patterns. These can be considered to be univariate time series of continuous values which are available at fine temporal scales. These time series data are long (with more and more data collected at finer resolutions), are asynchronous, with varying time lengths for different houses and sporadic missing values. Using probability distributions is a natural way to analyze this types of data because they are robust to uneven length, missing data, or noise. This paper proposes two approaches for obtaining pairwise similarities based on Jensen-Shannon distances between probability distributions across a selection of cyclic granularities. Cyclic temporal granularities [@Gupta2021-hakear], which are temporal deconstructions of a time period into units such as hour-of-the-day, work-day/weekend, can measure repetitive patterns in large univariate time series data. The resulting clusters are expected to group customers that have similar repetitive behaviors across cyclic granularities. The benefits of this approach are as follows.

- When using probability distributions, data does not have to be the same length or observed during the exact same time period (unless there is a structural pattern).

@@

- Jensen-Shannon distances evaluate the distance between two distributions rather than raw data, which is less sensitive to missing observations and outliers than other conventional distance methods.

- While most clustering algorithms produce clusters similar across just one temporal granularity, this technique takes a broader approach to the problem, attempting to group observations with similar distributions across all interesting cyclic granularities.

@@

<!--_Background and motivation_-->

The primary application of this work is data from the Smart Grid, Smart City (SGSC) project (2010–2014) available through the [Department of the Environment and Energy](https://data.gov.au/data/organization/doee). Half-hourly measurements of usage for more than 13,000 household electricity smart meters is provided from from October 2011 to March 2014. <!--Larger data sets include greater uncertainty about customer behavior due to growing variety of customers.--> Households vary in size, location, and amenities such as solar panels, central heating, and air conditioning. The behavioral patterns differ amongst customers due to many temporal dependencies. Some households use a dryer, while others dry their clothes on a line. Their weekly usage profile may reflect this. They may vary monthly, with some customers using more air conditioners or heaters than others, while having equivalent electrical equipment and weather circumstances. Some customers are night owls, while others are morning larks. Daily energy usage varies depending on whether customers stay home or work away from home. Age, lifestyle, family composition, building attributes, weather, availability of diverse electrical equipment, among other factors, make the task of properly segmenting customers into comparable energy behavior complex. The challenge is to be able to cluster consumers into these type of expected patterns, and other unexpected patterns, using only their energy usage history (@Ushakova2020-rl). <!--To safeguard the customers' privacy, it is probable that such information is not accessible. Also, energy suppliers may not always update client information, such as property features, in a timely manner.--> There is a growing need to have methods that can examine the energy usage heterogeneity observed in smart meter data and what are some of the most common power consumption patterns.

<!-- _Related work_-->

There is a growing body of literature focused on time series clustering related to smart meter data. @Tureczek2017-pb conducted a systematic study of over $2100$ peer-reviewed papers on smart meter data analytics. <!--None of the $34$ articles chosen for their emphasis use Australian smart meter data.--> The most often used algorithm is $k$-means [@Rhodes2014-it]. 
$k$-means can be made to perform better by explicitly incorporating time series features such as correlation or cyclic patterns rather than performing it on raw data. To reduce dimensionality, several studies use principal component analysis (PCA) or factor analysis to pre-process smart-meter data before clustering (@Ndiaye2011-pf). PCA eliminates correlation patterns and decreases feature space, but loses interpretability. Other algorithms utilized in the literature include $k$-means variants, hierarchical clustering, and greedy $k$-medoids. Time series data, such as smart meter data, are not well-suited to any of the techniques mentioned in @Tureczek2017-pb. Only one study [@ozawa2016determining] identified time series characteristics by first conducting a Fourier transformation, to convert data from time to frequency domain, followed by $k$-means to cluster by greatest frequency. @Motlagh2019-yj suggests that the time feature extraction is limited by the type of noisy, patchy, and unequal time-series common in residential customers and addresses model-based clustering by transforming the series into other objects such as structure or set of parameters which can be more easily characterized and clustered. [@chicco2010renyi] addresses information theory-based clustering such as Shannon or Renyi entropy and its variations. @Melnykov2013-sp discusses how outliers, noisy observations and scattered observations can complicate estimating mixture model parameters and hence the partitions. None of these methods focuses on exploring heterogeneity in repetitive patterns based on the dynamics of multiple temporal dependencies using probability distributions, which forms the basis of the methodology reported here.

@@

This paper is organized as follows.  Section~\ref{sec:methodology} provides the clustering methodology. Section~\ref{sec:validation} shows data designs to validate our methods. Section~\ref{sec:application-gracsr} discusses the application of the method to a subset of the real data. Finally, we summarize our results and discuss possible future directions in Section~\ref{sec:discussion-gracsr}.

@@

<!--The foundation of our method is unsupervised clustering algorithms based exclusively on the time-series data. The proposed methods leverage the intrinsic data structure hidden within cyclic temporal granularities.--> The existing work on clustering probability distributions assumes we have an independent and identically distributed samples $f_1(v),\dots,f_n(v)$, where $f_i(v)$ denotes the distribution from observation $i$ over some random variable $v = \{v_t: t = 0, 1, 2, \dots, T-1\}$ observed across $T$ time points. <!--So going back to the smart meter example, $f_i(v)$ is the distribution of customer $i$ and $v$ is electricity demand.--> In our approach, instead of considering the probability distributions of the linear time series, we compare them across different categories of a cyclic granularity. We can consider categories of an individual cyclic granularity ($A$) or combination of categories for two interacting granularities ($A*B$) to have a distribution, where $A, B$ are defined as $A = \{a_j: j=1, 2, \dots J\}$ and $B = \{b_k: k = 1, 2, \dots K\}$. For example, let us consider two cyclic granularities, $A$ and $B$, representing hour-of-day and day-of-week, repsectively. Then $A = \{0, 1, 2, \dots, 23\}$ and $B = \{Mon, Tue, Wed, \dots, Sun\}$. In case individual granularities are considered, there are $J = 24$ distributions of the form $f_{i,j}(v)$ or $K = 7$ distributions of the form $f_{i,k}(v)$ for each customer $i$. In case of interaction, $J*K=168$ distributions of the form $f_{i,j, k}(v)$ could be conceived for each customer $i$. Hence clustering these customers is equivalent to clustering these collections of conditional univariate probability distributions. Towards this goal, the next step is to decide how to measure distances between collections of univariate probability distributions. <!--There are multiple ways to measure similarities depending on the aim of the analysis. <!--This paper focuses on looking at the (dis) similarity between underlying distributions that may have resulted in different patterns across different cyclic temporal granularities, that eventually have resulted in the (dis) similarity between time series. --> Here, we describe two approaches for finding distances between time series. Both of these approaches may be useful in a practical context, and produce very different but equally useful customer groupings. The distances can be supplied to any usual clustering algorithm, including $k$-means or hierarchical clustering, to group observations into a smaller more homogeneous collection. <!--These clusters may be commonly associated with real-world data segmentation. However, since the data is unlabeled a priori, more information is required to corroborate this.--> The flow of the procedures is illustrated in Figure \ref{fig:flowchart}.

```{r flowchart, fig.cap = "Flow chart illustrating the pipeline for out method for clustering time series.", out.width="100%"}
knitr::include_graphics("img/flowchart.png")

@@

### Selecting granularities

[@Gupta2021-hakear] provides a method for determining the significance of a cyclic granularity, and a ranking of multiple cyclic granularities. (This extends to harmonies, pairs of granularities that might interact with each other.) We define "significant" granularities as those with significant distributional differences across at least one category. The reason for subsetting granularities in this way, is that, clustering algorithms perform badly in the presence of nuisance variables. Granularities that do not have some difference between categories are likely to be nuisance variables. It should be noted that all of the time series in a collection may not have the same set of significant granularities. This is the approach for managing this generate a subset ($S_c$) of significant granularities across a collection of time series:

 (a) Remove granularities from the comprehensive list that are not significant for any time series.

 (b) Select only the granularities that are significant for the majority of time series.

<!--There will be time series in both cases where one or a few selected granularities are boring. Even in that case, having this group of observations with no interesting patterns at a granularity that regularly discovers patterns may be valuable. If, on the other hand, the granularities under examination are truly important for a group of data, unique patterns may be identified while clustering them.-->

### Data transformation

The shape and scale of the distribution of the measured variable (e.g. energy usage) affects distance calculations. Skewed distributions need to be symmetrized. Scales of individuals needs to be standardized, because clustering is to select similar patterns, not magnitude of usage. (Organizing individuals based on magnitude can achieved simply by sorting on a statistic like the average value across time.) <!--Observations often have a somewhat skewed time series distribution and their ranges might vary greatly. Statistical transformations are employed to bring all of them to the same range or normalize each observation. This is important because we are not interested in trivial clusters that vary in magnitudes, but rather in uncovering comparable patterns of distributional differences between categories.--> For the JS-based approaches, two data transformation techniques are recommended, normal-quantile transform (NQT) and robust scaling (RS). NQT is used in @@Gupta2021-hakear prior sorting granularities. <!-- $wpd$, which is the foundation of wpd-based distances. XX wpd was not yet introduced-->

- RS: The normalized $i^{th}$ observation is denoted by $v_{norm} = \frac{v_t - q_{0.5}}{q_{0.75}-q_{0.25}}$, where $v_t$ is the actual value at the $t^{th}$ time point and $q_{0.25}$, $q_{0.5}$ and $q_{0.75}$ are the $25^{th}$, $50^{th}$ and $75^{th}$ percentile of the time series for the $i^{th}$ observation. Note that, $v_{norm}$ has zero mean and median, but otherwise the shape does not change.

- NQT: The raw data for all observations is individually transformed [@Krzysztofowicz1997-bv], so that the transformed data follows a standard normal distribution. NQT will symmetrize skewed distributions. <!--As a result, determining which raw distribution was used is difficult using the modified distribution.--> A drawback is that any multimodality will be concealed. This, this should be checked prior to applying NQT.

@@

### Data pre-preprocessing

Computationally in R, the data is assumed to be a "tsibble object" (@wang2020tsibble) equipped with an index variable representing inherent ordering from past to present and a key variable that defines observational units over time. The measured variable for an observation is a time-indexed sequence of values. This sequence, however, could be shown in several ways. A shuffle of the raw sequence may represent hourly consumption throughout a day, a week, or a year. Cyclic granularities can be expressed in terms of the index set in the tsibble data structure.

The data object will change when cyclic granularities are computed, as multiple observations will be categorized into levels of the granularity, thus inducing multiple probability distributions. Directly computing Jensen-Shannon distances between the entire probability distributions can be computationally intensive. Thus it is recommended that quantiles are used to characterize the probability distributions. In the final data object, each category of a cyclic granularity corresponds to a list of numbers which is composed of a few quantiles.

### Distance metrics

@@

The total (dis) similarity between each pair of customers is obtained by combining the distances between the collections of conditional distributions. This needs to be done in a way such that the resulting metric is a distance metric, and could be fed into the clustering algorithm. Two types of distance metrics are considered:

@@

@@

The sum of the distances between two observations $x$ and $y$ in terms of cyclic granularity $A$ is defined as $$S^A_{x,y} = \sum_{j} D_{x,y}(A)$$ (sum of distances between each category $j$ of cyclic granularity $A$) or $$S^{A*B}_{x,y} = \sum_j \sum_k D_{x,y}(A, B)$$ (sum of distances between each combination of categories $(j, k)$ of the harmony $(A, B)$). After determining the distance between two series in terms of one granularity, we must combine them to produce a distance based on all significant granularities. When combining distances from individual $L$ cyclic granularities $C_l$ with $n_l$ levels, $$S_{x, y} = \sum_lS^{C_l}_{x,y}/n_l$$ is employed, which is also a distance metric since it is the sum of JS distances. This approach is expected to yield groups, such that the variation in observations within each group is in magnitude rather than distributional pattern, while the variation between groups is only in distributional pattern across categories.

Compute weighted pairwise distances $wpd$ (@Gupta2021-hakear) for all considered granularities for all observations. $wpd$ is designed to capture the maximum variation in the measured variable explained by an individual cyclic granularity or their interaction and is estimated by the maximum pairwise distances between consecutive categories normalized by appropriate parameters. A higher value of $wpd$ indicates that some interesting patterns are expected, whereas a lower value would indicate otherwise.

Once we have chosen $wpd$ as a relevant feature for characterizing the distributions across one cyclic granularity, we have to decide how we combine differences between the multiple features (corresponding to multiple granularities) into a single number. The Euclidean distance between them is chosen, with the granularities acting as variables and $wpd$ representing the value under each variable. With this approach, we should expect the observations with similar $wpd$ values to be clustered together. Thus, this approach is useful for grouping observations that have a similar significance of patterns across different granularities. Similar significance does not imply a similar pattern, which is where this technique varies from JS-based distances, which detect differences in patterns across categories.

@@

#### Number of clusters

Different approaches to determining number of clusters are required for varied applications and research objectives. It is often beneficial to investigate numerous cluster statistics in order to arrive at a number of clusters, since clustering has a variety of objectives (e.g., between-cluster separation, within-cluster homogeneity, representation by centroids, etc.), which could be in conflict with one another. Some typical ways of determining the number of clusters in the literature (e.g., the gap statistic [@tibshirani2001estimating], average silhouette width [@rousseeuw1987silhouettes], Dunn index [@dunn1973fuzzy]) are inspired by defining and balancing "within-cluster homogeneity" and "between-cluster separation". The Cluster separation index ($sindex$) proposed by @Hennig2014-ah is used as a guide to select the number of clusters in Section \ref{sec:validation}. $sindex$ [@R-fpc] is a compromise between minimum cluster separation and average cluster separation and is less sensitive to a single or few ambiguous points. $sindex$ cannot be optimized over the number of clusters, since increasing the number of clusters reduces it. The number of clusters required is determined by the slope going from steep to shallow (an elbow).

.<!--determined by calculating the distance between each observation and the cluster to which it does not belong. Then, the least 10% of these distances are averaged to only evaluate places near the cluster edges.-->

#### Algorithm

@@

#### Characterization of clusters

Cluster characterization is an important element of cluster analysis. @Cook2007-qe provides several methods for characterizing clusters. _Parallel coordinate plots_ (@wegman1990hyperdimensional), _Scatterplot matrix_, _Displaying cluster statistics_ (@dasu2005grouping), _MDS_ (@borg2005modern), PCA, _t-SNE_(@R-tsne), _Tour_ (@wickham2011tourr) are some of the graphical approaches used in this study. A Parallel Coordinates Plot features parallel axes for each variable and each axis is linked by lines. Changing the axes may reveal patterns or relationships between variables for categorical variables. However, for categories with cyclic temporal granularities, preserving the underlying ordering of time is more desirable. Displaying cluster statistics is useful when we have larger problems and it is difficult to read the parallel coordinate plots due to congestion. All of MDS, PCA and t-SNE use a distance or dissimilarity matrix to construct a reduced-dimension space representation, but their goals are diverse. Multidimensional scaling (@borg2005modern) seeks to maintain the distances between pairs of data points, with an emphasis on pairings of distant points in the original space. The t-SNE embedding will compress data points that are close in high-dimensional space. Tour is a collection of interpolated linear projections of multivariate data into lower-dimensional space. The cluster characterization approach varies depending on the distance metric used. Parallel coordinate plots, scatter plot matrices, MDS or PCA are potentially useful ways to characterize clusters using wpd-based distances. For JS-based distances, plotting cluster statistics is beneficial for characterization and variable importance could be displayed through parallel coordinate plots.

@@

### Data generation

@@

_Scenario (S1): All granularities significant_

@@

Consider the instance where $g1$, $g2$, and $g3$ all contribute to design distinction. This means that each granularity will have significantly different patterns at least across one of the designs to be clustered. In Table \ref{tab:tab-dist-design} various distributions across categories are considered (top) which lead to different designs (bottom). Figure \ref{fig:plot-3gran-new} shows the simulated variable's linear (left) and cyclic (right) representations for each of these five designs. The structural difference in the time series variable is impossible to discern from the linear view, with all of them looking very similar. The shift in structure may be seen clearly in the distribution of cyclic granularities. The following scenarios use solely graphical displays across cyclic granularities to highlight distributional differences in categories.

_Scenario (S2): Few significant granularities_
